---
title: "DE Analysis and Standard Visualizations"
output:
  html_document:
    toc: yes
    code_folding: hide
    toc_float: yes
params: 
  max_fdr: 0.05
  min_lfc: 0.585
  min_rowsum: 10
  min_colsum: 20000
  indfilter: FALSE
  cookscutoff: FALSE
  countFile: '../feature_count/counts.s0.strict.txt'
  annoFile: 'https://raw.githubusercontent.com/radio1988/OneStopRNAseq/master/snakemake/workflow/resources/anno_tables/hg38.gencode.v34.te_included.anno.txt'
  metaFile: "../meta/meta.txt"
  contrastFile:  "../meta/contrast.de.csv"
  blackSamples: 'NA_AAAAXXskjk, NA_BBBXXXYjkdmf'
---

## Notes:
- Please search for “Warning” in this report for possible problems before proceeding
-	Normalized expression values, differentially expressed genes, and various types of plots such as heatmap and volcano plots are in the DESeq2 folder. For example, TPM.xlsx contains normalized expression values.
- Please use LFC_shrunken for Log2FoldChange, LFC_raw is not recommended
- Please use FDR (padj) instead of raw p-values for identifying significantly differentially expressed genes.
- blackSamples added in params
- manual dispersion estimation when DESeq() fails


```{r setup, include=F}
# Bioconductor
library(BiocManager)
library(DESeq2)
library(EnhancedVolcano)

# CRAN
library(tidyverse)
library(ggplot2)
library(ggpubr) # ggmaplot
library(ggrepel)
library(ashr)
library(MASS)
library(WriteXLS)
library(plyr)
library(gdata)
library(dplyr)
library(RColorBrewer)
library(pheatmap)
library(PoiClaClu)  
library(gridExtra)  
library(grid)
if(!require(berryFunctions)){
    install.packages("berryFunctions", repos = "http://cran.us.r-project.org")
    library(berryFunctions)
}

dir.create("./rnk", showWarnings = FALSE)
TEST=FALSE  # if TRUE, skip Excel outputs to save time
```

## Input parameter extraction
```{r params}
max_fdr <- params$max_fdr
min_lfc <- params$min_lfc
indfilter <- params$indfilter
cookscutoff <- params$cookscutoff

countFile <- params$countFile
annoFile <- params$annoFile
metaFile <- params$metaFile
contrastFile <-params$contrastFile

blackSamplesStr <- gsub(" " , '', params$blackSamples)
blackSamples <- unlist(strsplit(blackSamplesStr, ','))

paste("Data file name:", countFile)
paste("MetaData file name:", metaFile)
paste("Contrast file name:", contrastFile)
paste("Annotation file name:", annoFile)
paste("FDR cut-off:", max_fdr)
paste("Log2FC cut-off:", min_lfc)
cat('blackSamples to remove:', blackSamples)
```

## Functions
```{r functions}
fix_hyphen <- function(x){
  return(gsub("-", ".", x))
}

readTable <- function(fname){
  if (str_ends(fname, '.txt')){
      df <-  read_delim(fname, delim ="\t", comment = '#') 
   } else if (str_ends(fname, '.csv')){
     df <-  read_delim(fname, delim =",", comment = '#') 
   }else if (str_ends(fname, '.xlsx')){
      df <- readxl::read_xlsx( fname, na='NA', sheet=1)  # na term important
      df <- data.frame(df)  #important
      #df <- data.frame(lapply(df, fix_hyphen))
  } else{
      stop("file format not supported")
  }
  return (df)
}

get_density <- function(x, y, ...) {
  dens <- MASS::kde2d(x, y, ...)
  ix <- findInterval(x, dens$x)
  iy <- findInterval(y, dens$y)
  ii <- cbind(ix, iy)
  return(dens$z[ii])
}

volcanoplot <- function(res, anno, name='name', 
                        lfcthresh=2, sigthresh=0.05, 
                        labelsig=FALSE, xlim=100, ylim=1000, textcx=1) {
  # If Annotation file is wrong, No annotation match
  res0 <- res
  res<-merge(data.frame(res0), anno, by.x = 0, by.y=1, all.x=T, sort=F)
  if (sum(anno[, 1] %in% row.names(res0)) < 1) {
    warning(
      c("\nThe annotation file and the count filt does have no match in gene-id:\n", 
        "count table gene-id: ", head(row.names(res0), 1), 
        "\nanno table gene-id: ", anno[1:1, 1], 
        "gene-id rather than gene-name used for Volcano-plot"
      ))
    res$Name <- res[, 1]
  }
  
  # remove NA
  res$padj[is.na(res$padj)] <- 1
  # set lim on x, y
  res$padj[res$padj < 10^(-ylim) & !is.na(res$padj)] <- 10^(-ylim) # y-axis top value 50
  res$log2FoldChange[res$log2FoldChange > xlim] <- xlim
  res$log2FoldChange[res$log2FoldChange < -xlim] <- -xlim
  # show num_pos num_neg
  pos <- subset(res, padj<sigthresh & log2FoldChange>lfcthresh)
  neg <- subset(res, padj<sigthresh & log2FoldChange< -lfcthresh)
  pos.n <- dim(pos)[1]
  neg.n <- dim(neg)[1]
  
  EnhancedVolcano(res,
                       lab = res$Name,
                       #selectLab = as.character(res$Name[which(res$padj<labelcut)]), # mark top genes
                       #selectLab = c("FOS", "LDHA"), # mark selected genes
                       x = 'log2FoldChange',
                       y = 'padj',
                       title = name,
                       subtitle = paste("Up:", pos.n, ", Down:", neg.n, sep = ""),
                       xlab = bquote(~Log[2]~ "Fold Change"),
                       ylab = bquote(~-Log[10]~italic(FDR)),
                       pCutoff = max_fdr,
                       FCcutoff = min_lfc,
                       cutoffLineType = 'twodash',
                       cutoffLineWidth = 0.8,
                       legendLabels = c('NS', expression(Log[2]~FC),
                                        "FDR", expression(FDR~and~Log[2]~FC)),
                       caption = paste0('Total = ', nrow(res), ' genes'),
                       legendPosition = 'right',
                       legendLabSize = 10,
                       axisLabSize = 10,
                       legendIconSize = 3.0)
  ggsave(paste(name, "pdf", sep="."), width=8, height=6)
}

# Default plot
maplot <- function(res, anno, fname){
  maplotdat <- merge(anno, data.frame(res), by.x=1, by.y=0)
  ggmaplot(maplotdat, main = name,genenames = as.vector(maplotdat$Name),
           fdr = max_fdr, fc = 2^min_lfc, size = 0.4,
           xlab = "Log2 Mean Expression",
           ylab = "Log2 Fold Change",
           palette = c("#B31B21", "#1465AC", "darkgray"),
           legend = "top", top = 20,
           font.label = c("bold", 11),label.rectangle = TRUE,
           font.legend = "bold",
           font.main = "bold",
           ggtheme = ggplot2::theme_minimal())
  ggsave(fname)
}

zscore <- function(matrix){
  return( t(scale(t(matrix))))
}

rename_num_vector_by_order <- function(l){
  # l have to be a vector of numbers
  # output vector of roman numbers ordered by appearance in the input vector
  # e.g. c(2,3,3,2,1) -> c(I, II, II, I, III)
  # test rename_num_vector_by_order(c(2,3,3,2,1))
  u <- unique(l)
  n=0
  for (i in u){
    n = n+1; 
    l <- replace(l, l==i, as.character(as.roman(n)))
  }
  return(l)
}


Heatmap <- function(df, nclass=2, fname="heatmap", main="title"){
  if (dim(df)[1] > nclass){
    # Heatmap Pre-plot to get hclust
    my_sample_col <- data.frame(group = coldata[,'group'])
    row.names(my_sample_col) <- coldata[,'sample']
    p <- pheatmap(df, 
                  annotation_col = my_sample_col,
                  main = main,
                  cluster_cols = F, 
                  border_color = NA,
                  cutree_rows = nclass, 
                  breaks = seq(-3, 3, length.out = 100), 
                  show_rownames = F)
    p <- pheatmap(df, 
                  annotation_col = my_sample_col,
                  main = main,
                  cluster_cols = T, 
                  border_color = NA,
                  cutree_rows = nclass, 
                  breaks = seq(-3, 3, length.out = 100),                   
                  show_rownames = F)
    
    ## Print out gene classification (https://www.biostars.org/p/287512/)
    # cutree manually
    gene_classes <- sort(cutree(p$tree_row, k=nclass))
    gene_classes <- data.frame(gene_classes) 
    classification <- merge(gene_classes, df, by = 0)
    row.names(classification) <- classification$Row.names
    # Re-order original data (genes) to match ordering in heatmap (top-to-bottom)
    idx <- rownames(df[p$tree_row[["order"]],])
    classification <- classification[idx,] 
    # rename gene classes
    classification$gene_classes <- rename_num_vector_by_order(classification$gene_classes)
    # save excel
    if (!TEST){
      WriteXLS(classification, row.names = F,
               paste(fname,"gene_class.xlsx", sep = "."))}

    # get class label
    annotation_row = data.frame(
      class = classification$gene_classes
    )
    row.names(annotation_row) <- row.names(classification)
    annotation_row$class <- as.character(annotation_row$class)
    # output final plot
    if (dim(df)[1] < 20){ # manage row height
      p2 <- pheatmap(df,
                     annotation_col = my_sample_col,
                     main = main,
                     border_color = NA,
                     cluster_cols = F,
                     cutree_rows = nclass,
                     breaks = seq(-3, 3, length.out = 100),
                     show_rownames = F,
                     annotation_row = annotation_row,
                     cellheight = 20,
                     filename = paste(fname, "pdf", sep = "."))
      p3 <- pheatmap(df,  # sample clustering
                     annotation_col = my_sample_col,
                     main = main,
                     border_color = NA,
                     cluster_cols = T,
                     cutree_rows = nclass,
                     breaks = seq(-3, 3, length.out = 100),
                     show_rownames = F,
                     annotation_row = annotation_row,
                     cellheight = 20,
                     filename = paste(fname, "v2.pdf", sep = "."))
    }else{
      p2 <- pheatmap(df,
                     annotation_col = my_sample_col,
                     main = main,
                     border_color = NA,
                     cluster_cols = F,
                     cutree_rows = nclass,
                     show_rownames = F,
                     breaks = seq(-3, 3, length.out = 100),
                     annotation_row = annotation_row,
                     filename = paste(fname, "pdf", sep = "."))
      p3 <- pheatmap(df, # sample clustering
                     annotation_col = my_sample_col,
                     main = main,
                     border_color = NA,
                     cluster_cols = T,
                     cutree_rows = nclass,
                     show_rownames = F,
                     breaks = seq(-3, 3, length.out = 100),
                     annotation_row = annotation_row,
                     filename = paste(fname, "v2.pdf", sep = "."))
    }

  }else if (dim(df)[1] > 0){
    # output final plot
    if (dim(df)[1] < 20){ # manage row height
      p2 <- pheatmap(df,
                     annotation_col = my_sample_col,
                     main = main,
                     cluster_cols = F,
                     cluster_rows = F,
                     border_color = NA,
                     show_rownames = F,
                     breaks = seq(-3, 3, length.out = 100),
                     cellheight = 20,
                     filename = paste(fname, "pdf", sep = "."))
      p3 <- pheatmap(df,
                     annotation_col = my_sample_col,
                     main = main,
                     cluster_cols = T,
                     cluster_rows = F,
                     border_color = NA,
                     show_rownames = F,
                     breaks = seq(-3, 3, length.out = 100),
                     cellheight = 20,
                     filename = paste(fname, "v2.pdf", sep = "."))
    }else{
      p2 <- pheatmap(df,
                     annotation_col = my_sample_col,
                     main = main,
                     cluster_cols = F,
                     cluster_rows = F,
                     border_color = NA,
                     breaks = seq(-3, 3, length.out = 100),
                     show_rownames = F,
                     filename = paste(fname, "pdf", sep = "."))
      p3 <- pheatmap(df,
                     annotation_col = my_sample_col,
                     main = main,
                     cluster_cols = T,
                     cluster_rows = F,
                     border_color = NA,
                     breaks = seq(-3, 3, length.out = 100),
                     show_rownames = F,
                     filename = paste(fname, "v2.pdf", sep = "."))
    }
  }else{ print("no sig DEG for heatmap")}
}



process_deseq_res <- function(res="lfcshrink.res", res2="results.res", name='name', anno='anno.df', norm_exp="tpm.df"){
    ## Summary
    print(name)
    print("\n>>> Summary using FDR cut-off only (LFC not used)")
    summary(res, alpha=max_fdr)
    
    print("\n>>> Summary using both FDR and LFC_shrunken cut-off")
    sig_idx <- res$padj<max_fdr & abs(res$log2FoldChange) > min_lfc
    sig_idx[is.na(sig_idx)] <- FALSE
    res_sig <- res[sig_idx,]
    print(table(sig_idx))
    
    up_idx <- res$padj<max_fdr & res$log2FoldChange > min_lfc
    up_idx[is.na(up_idx)] <- FALSE
    res_sig <- res[up_idx,]
    print(table(up_idx))
    
    down_idx <- res$padj<max_fdr & res$log2FoldChange < -min_lfc
    down_idx[is.na(down_idx)] <- FALSE
    res_sig <- res[down_idx,]
    print(table(down_idx))

    # Prep
    res.df <- as.data.frame(res)
    names(res.df)[2] <- "log2FoldChange_shrunken"
    names(res.df)[3] <- "lfcSE_shrunken"

    res2.df <- as.data.frame(res2)
    names(res2.df)[2] <- "log2FoldChange_raw"
    names(res2.df)[3] <- "lfcSE_raw"
    res2.df <- res2.df[, c(2,3)]

    resdata <- merge(res.df, res2.df, by=0, sort=F, all.x=T)
    resdata <- merge(anno, resdata, by.x=1, by.y=1, sort=F, all.y=T)
    resdata <- merge(resdata, norm_exp, by.x=1, by.y=0, all.x=T, sort=F)
    head(resdata)
    sig_idx <- resdata$padj<max_fdr & abs(resdata$log2FoldChange_shrunken) > min_lfc # important to put this line right before output sig.xlsx
    sig_idx[is.na(sig_idx)] <- FALSE
    resdata.sig <- resdata[sig_idx,]
    head(resdata.sig)

    ## Write results
    if (!TEST){
      WriteXLS(x = resdata,
               ExcelFileName = paste(name, 'deseq2.xlsx', sep = '.'),
               row.names = F, SheetNames = 'sheet1', na = 'NA')  # for user
      
      WriteXLS(x = resdata.sig,
               ExcelFileName = paste(name, 'deseq2.sig.FDR', max_fdr, 
                                     'LFC', min_lfc, 'xlsx', sep = '.'),
               row.names = F, SheetNames = 'sheet1', na = 'NA')  # for user
    }

    ## For GSEA
    rnk <- subset(resdata, select = c("Name","log2FoldChange_shrunken"))
    colnames(rnk) <- c("# Name","log2FoldChange_shrunken")
    rnk <- rnk[order(rnk$log2FoldChange_shrunken), ]
    rnk[, 1] <- toupper(rnk[, 1])
    write.table(rnk, 
                paste('rnk/', name, '.rnk', sep = ''), 
                row.names = F, quote = F, sep='\t')


    # # Corrlelation of Length and LFC
    # resdata.sig.cor <- cor.test(resdata.sig$Length, 
    #                        resdata.sig$log2FoldChange_shrunken, 
    #                        method = "spearman")
    # title <- paste("Spearman Cor:", format(resdata.sig.cor$estimate, digits=2, nsmall=2),
    #                "p-value:", format(resdata.sig.cor$p.value, digits=3, nsmall=3),
    #                sep = " ")
    # 
    # resdata.sig$density <- get_density(resdata.sig$Length, resdata.sig$log2FoldChange_shrunken, n = 100)
    # # set lim (careful, only after outputting excel)
    # resdata.sig$log2FoldChange_shrunken[resdata.sig$log2FoldChange_shrunken > 10] <- 10
    # resdata.sig$log2FoldChange_shrunken[resdata.sig$log2FoldChange_shrunken < -10] <- -10
    # resdata.sig$Length[resdata.sig$Length > 20000] <- 20000
    # ggplot(resdata.sig) + 
    #     geom_point(aes(Length, log2FoldChange_shrunken, color = density)) +
    #     scale_color_viridis() +
    #     ggtitle(paste(paste("Sig-DEG for", name), title, sep = "\n") ) + 
    #     ylim(-10, 10) +        
    #     xlim(0,20000)
    
    ##  Plots
    ggplot(data.frame(res), aes(x=pvalue))+
      geom_histogram(color="darkblue", fill="lightblue")
    ggsave(paste0(name, '.pvalue.pdf'))
    
    ggplot(data.frame(res), aes(x=padj))+
      geom_histogram(color="darkblue", fill="lightblue")
    ggsave(paste0(name, '.fdr.pdf'))
    
    maplot(res, anno, paste0(name, '.maplot.shrunken_lfc.pdf'))
    maplot(res2, anno, paste0(name, '.maplot.raw_lfc.pdf'))
    
    volcanoplot(res, anno, lfcthresh=min_lfc, sigthresh=max_fdr,
                textcx=.8,  name= paste(name, "LFC_shrunken", sep="."))
    volcanoplot(res2, anno, lfcthresh=min_lfc, sigthresh=max_fdr,
                textcx=.8,name= paste(name, "LFC_raw", sep="."))
    
    n1 <- dim(resdata.sig)[2]
    n2 <- dim(norm_exp)[2]
    zscore.df <- zscore(resdata.sig[, (n1-n2+1):n1])
    rownames(zscore.df) <- resdata.sig[,1]
    colnames(zscore.df) <- gsub (":TPM", "", colnames(zscore.df))
    colnames(zscore.df) <- gsub (":DESeq2NormalizedCount", "", colnames(zscore.df))
    Heatmap(zscore.df, nclass = 2,
            fname = paste(name, "heatmap", sep="."),
            main = paste(name, "LFC >", min_lfc, "FDR <", max_fdr ))
}
```

## Importing data
```{r read_meta_data}
meta.df <- readTable(metaFile)
meta.df <- meta.df[!(meta.df[, 1] %in% blackSamples), ]

contrast.df <- readTable(contrastFile)
print(contrast.df)
```

```{r read_count_data}
# read
paste('countFile:', countFile)
if (grepl("txt$", countFile)){
  df <- read_delim(countFile, delim ="\t", comment = '#') 
  df <- column_to_rownames(df, 'Geneid')
} else if (grepl("xlsx$", countFile)){
  df <- readTable(countFile)  
} else {
  stop("input count file not xlsx, nor txt format, abort!!!")
}

# auto-detect content format
featureCountsFormat <- F
cleanCountFormat <- F
osrCountFormat <- F

if (colnames(df)[1] == "Chr" & colnames(df)[2] == "Start" & colnames(df)[3] == "End"){
  cat("featureCounts format table detected\n")
  featureCountsFormat <- T
  # clean names
  colnames(df) <- gsub("\\.bam$", "", colnames(df))
  colnames(df) <- gsub("sorted_reads.", "", colnames(df))
  colnames(df) <- gsub("mapped_reads.", "", colnames(df))
  # convert to int
  df[, 6:ncol(df)] <- sapply(df[, 6:ncol(df)], as.integer)
  # get cts
  cts <- as.matrix(df[, 6:ncol(df)])
} else if (dim(df)[2] == dim(meta.df)[1] + 1){
  cat("clean format table detected\n")
  cleanCountFormat<- T
  row.names(df) <- df[, 1]
  df[, 2:ncol(df)] <- sapply(df[, 2:ncol(df)], as.integer)
  cts <- as.matrix(df[, 2:ncol(df)])
} else if (colnames(df)[1] == "Gene" & colnames(df)[2] == "Name" & colnames(df)[3] == "Type") {
  cat("OSR COUNT.xlsx format detected\n")
  osrCountFormat <- T
  # clean names
  colnames(df) <- gsub(".COUNT", "", colnames(df))
  # convert to int
  df[, 4:ncol(df)] <- sapply(df[, 4:ncol(df)], as.integer)
  row.names(df) <- df$Gene
  # get cts
  cts <- as.matrix(df[, 4:ncol(df)])
} else {
  stop("COUNT file not in featureCounts output format, nor cleanCountFormat format, nor osrCount format. \nMay have different number of samples in meta-data and count-table if cleanCountFormat is provided\n")
}

# remove blackSamples
df <- df[, !(colnames(df) %in% blackSamples)]
cts <- cts[, !(colnames(cts) %in% blackSamples)]

# check sample names
if (all(meta.df[[1]] %in% colnames(cts)) ){
  cat("all samples in meta.df included in count table")
} else {
  warning("count-table: ", colnames(cts), "\n")
  warning("meta-data: ",meta.df[[1]], "\n")
  stop("some samples in meta not included in count table")
}

# match order
cts_idx <- match(meta.df[[1]], colnames(cts))
cts <- cts[,cts_idx]

# Report Summary
print(paste("lib-size in millions:"))
print(format(colSums(cts/1e6), digits=2))
print(paste("Dim of input data:"))
print(dim(cts))
```

## Importing annotation
```{r import_annotation}
# from GitHub (must be raw, not zipped)
getAnnotation <- function(urlpath) {
  tmp <- tempfile()
  download.file(urlpath, destfile = tmp, method = 'auto')
  return(read.table(tmp, sep="\t", header = TRUE, quote=""))
}

paste("Getting data from", annoFile)

if (grepl('https://', annoFile)){
  print("downloading remote annoFile")
  anno <- getAnnotation(annoFile)
}else{
  print("reading local annoFile")
  anno <- read.table(annoFile, sep = "\t", header = T, quote="")
}
anno <- anno[!duplicated(anno[1]), ] # assume first column is Gene ID
print("Dimention of annotation table: ")
dim(anno)
head(anno)

if (sum(anno[, 1] %in% row.names(cts)) < 1){
  warning("!!! Annotation file and count file have no ID in common")
  warning("The results will be unannotated")
  warning("Please Double check Annotation file")
  print("count table ID:")
  print(row.names(cts)[1:2])
  print("anno table ID:")
  print(anno[1:2, 1])
  }
```

## COUNT output (without filtering)
```{r count}
count <- data.frame(cts)
colnames(count) <- paste(colnames(count),"COUNT", sep = ":")
count_out <- merge(anno, count, by.x=1, by.y=0, all.y=T, sort=F)
head(count_out)
if (!TEST){
  WriteXLS(x = count_out, ExcelFileName = 'COUNT.xlsx', row.names = F, SheetNames = 'sheet1', na = 'NA')
  print("saved in COUNT.xlsx")}
```

## Filtering
```{r filtering}
min_rowsum <- params$min_rowsum
gene_filter <- rowSums(cts) >= min_rowsum  # default 10

min_colsum <- params$min_colsum
sample_filter <- colSums(cts) >= min_colsum

if (sum(gene_filter) > 2 & sum(sample_filter) >= 4){
  kept_samples <-  names(sample_filter)[sample_filter]
  meta.df <- meta.df[meta.df[[1]] %in% kept_samples,]
  group_count <- plyr::count(meta.df[[2]])
  cts <- cts[gene_filter, sample_filter]
  if (featureCountsFormat) {
    df <- df[gene_filter, c(rep(TRUE,5), sample_filter )]
  }else if (cleanCountFormat) {
    df <- df[gene_filter, sample_filter]
  } else if (osrCountFormat) {
    df <- df[gene_filter, c(rep(TRUE,3), sample_filter )]
  }    
  
  # match order
  meta.df <- meta.df[match(colnames(cts), meta.df[[1]]), ]
  cat(paste("Removed genes with less than ", min_rowsum, "reads/fragments across all samples", "\n"))
  cat(paste("Removed samples with less than ", min_colsum, "reads/fragments across all genes", "\n"))
  cat(paste("Data dim after filtering:"))
  dim(cts)
  print(group_count)
  boxplot(
            log10(cts+1), 
            las=2, 
            main = "Raw Gene Count Distribution After Filtering",
            ylab = 'log10(count+1)'
        )
}else{
  print("WARNING!!!: Library size too small, too few genes/samples would be left after filtering, so skipped filtering.")
  print("PLEASE interpret results with caution!!!")
  file.create('WARNING.LowCountSamples.Caution')
  sample_filter[1:length(sample_filter)] <- rep(TRUE, length(sample_filter))
  gene_filter[1:length(gene_filter)] <- rep(TRUE, length(gene_filter))
  kept_samples <-  names(sample_filter)[sample_filter]
  meta.df <- meta.df[meta.df[[1]] %in% kept_samples,]
  group_count <- plyr::count(meta.df[[2]])
  cts <- cts[gene_filter, sample_filter]
  if (featureCountsFormat) {
    df <- df[gene_filter, c(rep(TRUE,5), sample_filter )]
  }else if (cleanCountFormat) {
    df <- df[gene_filter, sample_filter]
  } else if (osrCountFormat) {
    df <- df[gene_filter, c(rep(TRUE,3), sample_filter )]
  }    
  
  # match order
  meta.df <- meta.df[match(colnames(cts), meta.df[[1]]), ]
  cat(paste("Removed genes with less than ", min_rowsum, "reads/fragments across all samples", "\n"))
  cat(paste("Removed samples with less than ", min_colsum, "reads/fragments across all genes", "\n"))
  cat(paste("Data dim after filtering:"))
  dim(cts)
  print(group_count)
  boxplot(
            log10(cts+1), 
            las=2, 
            main = "Raw Gene Count Distribution After Filtering",
            ylab = 'log10(count+1)'
        )
}
```

## COUNT calculation again (after filtering)
```{r filtered_count}
count <- data.frame(cts)
colnames(count) <- paste(colnames(count),"COUNT", sep = ":")
count_out <- merge(anno, count, by.x=1, by.y=0, all.y=T, sort=F)
head(count_out)
```

## TPM calculation
```{r tpm}
calculateTPM <- function(counts,len) {
  # michael's version
  # https://support.bioconductor.org/p/91218/
  x <- counts/len
  return(t(t(x)*1e6/colSums(x)))
}

if (featureCountsFormat ){
  tpm <- calculateTPM(cts, df$Length)
  tpm <- data.frame(tpm)
  colnames(tpm) <- paste(colnames(tpm),"TPM",  sep = ":")
  tpm_out <- merge(anno, tpm, by.x=1, by.y=0, all.y=T, sort=F)
  # head(tpm_out)
  # tail(tpm_out)
  WriteXLS(x = tpm_out, 
           ExcelFileName = 'TPM.xlsx', row.names = F, SheetNames = 'sheet1', na = 'NA')
  print("saved in TPM.xlsx")
} else {
  cat ("TPM not calculated if COUNT-table is not featureCount format\n")
}
```

## FPKM calculation
```{r fpkm}
calculateFPKM <- function(counts,len) {
  x <- counts
  x <- t(t(x)*1e6/colSums(x))
  return (x/len*1e3)
}

if (featureCountsFormat ){
  fpkm <- calculateFPKM(cts, df$Length)
  fpkm <- data.frame(fpkm)
  colnames(fpkm) <- paste(colnames(fpkm), "FPKM", sep = ":")
  fpkm_out <- merge(anno, fpkm, by.x=1, by.y=0, all.y=T, sort=F)
  # head(fpkm_out)
  # tail(fpkm_out)
  if (!TEST){
    WriteXLS(x = fpkm_out, 
             ExcelFileName = 'FPKM.xlsx', row.names = F, SheetNames = 'sheet1', na = 'NA')
    print("saved in FPKM.xlsx")
    print("Recommend to use TPM, rather than FPKM")}
} else {
  cat ("FPKM not calculated if COUNT-table is not featureCount format\n")
}
```

## Design matrix extracted from meta-data
```{r coldata}
meta_idx <- match(colnames(cts), meta.df[[1]])
meta_idx <- meta_idx[!is.na(meta_idx)]
meta.df <- meta.df[meta_idx, ]
cts <- cts[, colnames(cts) %in% meta.df[[1]]]


sample <- factor(meta.df[[1]])
batch <- factor(meta.df[[3]])
group <- factor(meta.df[[2]])

coldata <- data.frame(row.names=colnames(cts), 
                      sample,
                      group,
                      batch
                      )
coldata
```

```{r SaveWorkSpace}
save.image()
```

## Model fitting
```{r dds}
if (length(levels(batch)) > 1){
  dds <- DESeqDataSetFromMatrix(countData = cts, 
                                colData = coldata, 
                                design = ~  0 + group + batch)
}else{
  dds <- DESeqDataSetFromMatrix(countData = cts, 
                                colData = coldata, 
                                design = ~  0 + group)  # converted to alph-order
}
dds
if (is.error(dds <-DESeq(dds))){
  cat("using manual dispersion estimation")
  dds <- estimateSizeFactors(dds)
  dds <- estimateDispersionsGeneEst(dds)
  dispersions(dds) <- mcols(dds)$dispGeneEst
  dds <- nbinomWaldTest(dds)
}
resultsNames(dds)
#saveRDS(dds, file = 'deseq2.dds.rds')
```

## Save DESeq2 normalized counts
```{r deseq2_norm}
normalized_counts <- counts(dds, normalized=TRUE)
normalized_counts <- data.frame(normalized_counts)

boxplot(
          log10(normalized_counts + 1), 
          las=2, 
          main = "Normazlied Gene Count Distribution",
          ylab = 'log10(norm+1)'
      )
  
colnames(normalized_counts) <- paste(colnames(normalized_counts),"DESeq2NormalizedCount",  sep = ":")
normalized_counts_out <- merge(anno, normalized_counts, by.x=1, by.y=0, all.y=T, sort=F)
if (!TEST){
  WriteXLS(x = normalized_counts_out, 
           ExcelFileName = 'DESeq2NormalizedCounts.xlsx', row.names = F, SheetNames = 'sheet1', na = 'NA')
  print("saved in DESeq2NormalizedCounts.xlsx")
}
```

<!-- ## QC Plots -->
<!-- ## Data transformation -->
<!-- ```{r} -->
<!-- #vsd <- vst(dds, blind=FALSE) -->
<!-- rld <- rlog(dds, blind=FALSE) -->
<!-- counts <- counts(dds, normalized=0) -->
<!-- logCounts <- log10(counts +1 ) -->
<!-- normed <- counts(dds, normalized=1) -->
<!-- logNormed <- log10(normed+1) -->
<!-- ``` -->

### Histogram of Log10(Counts)
```{r histogram}
log1p_count <- log10(counts(dds) + 1)
hist(log1p_count, 
     main = 'Histogram of log10(count + 1)', 
     xlab = "log10(count+1)",
     100) # by default, use non-normalized data by counts function
```

### Dispersion plot
```{r dispersion_plot}
plotDispEsts(dds, main="Dispersion plot")
```

### Sample PCA plot
```{r pca}
plotQC_PCA <- function(dds, fname='pca.pdf') {
  vsd <- varianceStabilizingTransformation(dds) # fixed num < num(rowsum>5)
  pcaData <- plotPCA(vsd, intgroup = 'group', returnData=TRUE) # labeling fixed
  percentVar <- round(100 * attr(pcaData, 'percentVar'), 1)
  if (length(levels(batch)) > 1){
    ggplot(pcaData, aes(PC1, PC2, color = group, shape = batch)) +
    geom_point(size = 3) +
    xlab(paste0("PC1: ", percentVar[1], "% variance")) +
    ylab(paste0("PC2: ", percentVar[2], "% variance")) +
    geom_label_repel(aes(label = sample),
                      box.padding = 0.35,
                      point.padding = 1,
                      segment.color = 'grey50',
                      segment.alpha = 0.5,
                      show.legend = FALSE) + # if TRUE, legend display might not be correct
    theme_classic()
    ggsave(fname)

  }else{
    ggplot(pcaData, aes(PC1, PC2, color = group)) +
    geom_point(size = 3) +
    xlab(paste0("PC1: ", percentVar[1], "% variance")) +
    ylab(paste0("PC2: ", percentVar[2], "% variance")) +
    geom_label_repel(aes(label = sample),
                      box.padding = 0.35,
                      point.padding = 1,
                      segment.color = 'grey50',
                      segment.alpha = 0.5,
                      max.overlaps = 20,
                      show.legend = FALSE) + # if TRUE, legend display might not be correct
    theme_classic()
    ggsave(fname)
  }
}

plotQC_PCA_no_label <- function(dds, fname='pca.pdf') {
  vsd <- varianceStabilizingTransformation(dds) # fixed num < num(rowsum>5)
  pcaData <- plotPCA(vsd, intgroup = 'group', returnData=TRUE) # labeling fixed
  percentVar <- round(100 * attr(pcaData, 'percentVar'), 1)
    if (length(levels(batch)) > 1){
        ggplot(pcaData, aes(PC1, PC2, color = group, shape = batch)) +
        geom_point(size = 3) +
        xlab(paste0("PC1: ", percentVar[1], "% variance")) +
        ylab(paste0("PC2: ", percentVar[2], "% variance")) +
        theme_classic()
        ggsave(fname)
    }else{
        ggplot(pcaData, aes(PC1, PC2, color = group)) +
        geom_point(size = 3) +
        xlab(paste0("PC1: ", percentVar[1], "% variance")) +
        ylab(paste0("PC2: ", percentVar[2], "% variance")) +
        theme_classic()
        ggsave(fname)
    }
}

plotQC_PCA_no_label(dds, "sample_PCA.pdf")

if (max(unlist(lapply(meta.df[[1]], nchar))) <= 36){
  plotQC_PCA(dds, "sample_PCA.labeled.pdf")
}
```

### Sample heatmap (Poisson Distance Based)
```{r poisson_heatmap}
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
poisd <- PoissonDistance(t(counts(dds)))
samplePoisDistMatrix <- as.matrix( poisd$dd ) 
rownames(samplePoisDistMatrix) <- coldata$sample
colnames(samplePoisDistMatrix) <- NULL 
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)

my_sample_col <- data.frame(group = coldata[,'group'])
row.names(my_sample_col) <- coldata[,'sample']
pheatmap(samplePoisDistMatrix,
         annotation_row = my_sample_col,
         clustering_distance_rows=poisd$dd,
         clustering_distance_cols=poisd$dd,
         col=colors, 
         clustering_method='complete', 
         legend = T, 
         filename = "sample_poisson_distance.pdf")
```

## DE analysis
```{r, warning=T, error=T, DE_ANALYSIS}
parse_name <- function(name){
  name <- gsub(" ", "", name)
  name <- gsub(";$", "", name)
  names <- strsplit(name, ";") [[1]]
  name <- gsub(";", ".", name)
  return (list(name=name, names=names))
}

for (i in 1:dim(contrast.df)[2]){
  # parse 
  name1 <- parse_name(contrast.df[1,i])
  name2 <- parse_name(contrast.df[2, i])
  name <- paste(name1$name, name2$name, sep = "_vs_")
  if (nchar(name) > 100) {name = paste0('contrast', i)}
  print(paste(">>>", i, name))
  
  # check group_count for all groups in each contrast
  groups <- unique(c(name1$names, name2$names))
  
  with_sample <- groups %in% group_count[[1]]  # count >= 1
  if (!all(with_sample)){
    print(paste('some groups in', name, 'has no filtered samples, skipped !!!'))
    print(group_count)
    next
    }

  group_with_reps <-  group_count[[1]][group_count$freq >= 2]
  
  with_reps1 <- any(name1$names %in% group_with_reps)
  with_reps2 <- any(name2$names %in% group_with_reps)
  
  if (!any(with_reps1, with_reps2)) {
    print(paste('no groups in', name, 'has reps, skipped !!!'))
    print(group_count)
    next
  }

  # analysis
  gsub("group", "", resultsNames(dds))
  poss <- match(name1$names, gsub("group", "", resultsNames(dds)))
  negs <- match(name2$names, gsub("group", "", resultsNames(dds)))
  contrast <- rep(0, length(resultsNames(dds)))
  contrast[poss] <- 1/length(poss)
  contrast[negs] <- -1/length(negs)
  print(data.frame(resNames=gsub("group", "", resultsNames(dds)), 
                   contrast=contrast))
  
  res <- lfcShrink(dds, contrast = contrast, type = 'ashr')
  res2 <- results(dds, contrast = contrast, 
                  independentFilter=indfilter, cooksCutoff=cookscutoff)
  if (featureCountsFormat){
    process_deseq_res(res = res, res2=res2, name = name, anno = anno, norm_exp = tpm)
  } else if (cleanCountFormat | osrCountFormat){
      process_deseq_res(res = res, res2=res2, name = name, anno = anno, norm_exp = normalized_counts)
  } else {
    stop("COUNT table format was wrong")
  }
} 
```

## Log
```{r log}
save.image()
sessionInfo()
```

